---
title: Meta-Analsis for the Binomial Model with Normal Random Effects
subtitle: Heirarchical and Empirical Bayesian Approaches
author: Sophie Huebler
format: revealjs
embed-resources: true
editor: 
  markdown: 
    wrap: 72
---

# Section 1: Introduction to the Problem

## The Goal

1)  Create a computationally efficient implementation of the
    gibbs-metropolis-hastings hybrid algorithm for a heirarchical
    bayesian approach to meta analysis.

2)  Create a computationally efficient function to compute mariginal
    likelihood and optimize it to empirically estimate parameters for
    random effects for an empirical bayesian approach to meta analysis.

## Meta-Analysis {.scrollable .smaller}

We will use a classic meta-analysis case to motivate this problem. Our
treatment effect of interest is the odds ratio of an event occurring
between treatment and control groups, and there are 7 studies which have
estimated this effect by recording the number of events and sample size
in a treatment sample and a control sample.

```{r}
#| label: some-code
library(MASS)
library(epiworldR)
library(tidyverse)
library(metafor)
```

```{r}
source("data/read_data.R")
```

```{r}
dat |>
  dplyr::select(trial:nic)|>
  kableExtra::kbl(digits = 2,
      caption = "Data",
      col.names = c("Trial",
                    "R_it", "N_it",
                    "R_ic", "N_ic"))|>
  kableExtra::kable_classic_2(html_font = "Cambria")
```

## Meta-Analysis {.scrollable .smaller}

Each study therefore has a log odds ratio $\theta_i$ which estimates the
study specific treatment effect, and a standard error for that estimate.
We assume that due to differing study protocol and random chance, each
study is estimating a log odds ratio that is drawn from a normal
distribution with center $\mu$ and variance $\tau^2$. We will call this
the random effect population prior $\pi(\theta_i | \mu, \tau)$.

```{r}
dat |>
  dplyr::select(trial:v_i)|>
  kableExtra::kbl(digits = 2,
      caption = "Data",
      col.names = c("Trial",
                    "R_it", "N_it",
                    "R_ic", "N_ic",
                    "log(OR)", "Var(log(OR))"))|>
  kableExtra::kable_classic_2(html_font = "Cambria")
```

## The Model

$$
\begin{aligned}
&i \in i,…,k \hspace{0.5cm} and \hspace{0.5cm} j \in \{C,T\}\\
& Y_{ij}| \pi_{ij},n_{ij} \sim Bin(n_{ij}, \pi_{ij})\\
&logit \pi_i^C | \gamma_i, \theta_i = \gamma_i - \theta_i/2\\
&logit \pi_i^T | \gamma_i, \theta_i = \gamma_i + \theta_i/2\\
& logit(\frac{\pi_{iT}}{\pi_{iC}}) = \theta_i | \mu, \tau^2 \sim N(\mu, \tau^2)\\
& \mu \sim N(a_1, b_1 )\\
& 1/\tau^2\sim gamma(a_2, b_2)\end{aligned}
$$

## Heirarichal Estimation

From the model, we can easily write out the joint distribution and the
full conditionals for $\mu$, $\tau^2$, $\bf{\theta}$, and $\bf{\gamma}$
(I leave this as an exercise to the listener). This allows us to use the
gibbs sampling procedure to iteratively update posteriors. It is however
important to note that the thetas and gammas do not have closed form
full conditionals. Therefore we must also incorporate a metropolis
hastings component wise updating procedure.

## Empirical Estimation

Instead of iteratively updating the $\mu$ and $\tau^2$ variables, we can
estimate them from the marginal likelihood using an MLE approach after
integrating out all other parameters, thus reducing the problem to a
single level model.

# Section 2: Solution Plan

## Heirarchical Bayesian Approach

-   Full conditionals:

    -    Write in both R and c++

-   Wrapper for the full algorithm:

    -   Rewrite with data table so that storage and manipulation is
        easier

    -    Pre-allocate memory

-    Use microbenchmark to comapre efficiency to original code

## Empirical Bayesian Approach

-   Marignal likelihood

    -   Use parallel programming for the R implementation that currently
        simulates average likelihood of the marginal likelihood

    -    Write the full likelihood function in C++ and then use GNU
        scientific library for numeric integration

# Section 3: Preliminary Results

## Heirarchical Bayesian {.scrollable .smaller}

Currently all posterior functions are written in modularized format in R
and wrapped in a function that stores the results as a data table.

-   Old function: 2 for loops, 4 lapply statements

-   New function: 1 for loop

```{r}
source("Functions/Original_Code.R")
source("Functions/set_hyperparameters.R")
source("Functions/post_mu.R")
source("Functions/post_tau2.R")
source("Functions/post_gamma.R")
source("Functions/post_theta.R")
source("Functions/mh.R")
source("Functions/gibbs.R")
source("Functions/gibbs_mh_run.R")
```

```{r}
hypers <- set_hyperparameters()
```

## Marginal Likelihood {.scrollable .smaller}

Original code:

```{r, echo = T}
mlik_func <- function(mu,
                      tau2,
                      Y, 
                      #simulate
                      sim = 1000
                      ){
  
#Set up
  #number of studies
  k = nrow(Y)
  #initialize 
  lik <- 0 
  #test <-  0
  
#Outer loop !!! This is where parallelization can be implemented
  for(j in 2:sim){
    #Simulate theta and gamma
    #based on mu and tau2 to optimize and take average lik
    theta <- rnorm(k, mean = mu, sd = sqrt(tau2))
    gamma <- rnorm(k, mean = 0, sd = 10)
    
    #Calculate odds (expit scale)
    eta_C <- gamma - theta/2
    eta_T <- gamma + theta/2
    
    #Transform odds to probabilities (p scale)
    p_C <- 1/(1+exp(-eta_C))
    p_T <- 1/(1+exp(-eta_T))
    
    #Construct priors (on theta and gamma only)
    #Use log scale for easier computation so
    #addition instead of multiplication for
    #likelihood and priors
    
    pi_theta <- dnorm(theta, mean = mu, sd = sqrt(tau2), log = T)
    pi_gamma <- dnorm(gamma, mean = 0, sd = 10, log = T)
    
    #Construct likelihood for both control
    #and treatments, because the likelihood 
    #depends on the data (do not construct
    #likelihood for the parameters being estimated
    #or integrated out)
    
    #Use probabilities found above
    
    lik_C <- dbinom(Y[,"ric"], Y[,"nic"], p_C, log = T)
    lik_T <- dbinom(Y[,"rit"], Y[,"nit"], p_T, log = T)
    
    #Need to tranform back to exp scale
    lik<- lik + exp(lik_C+lik_T+pi_theta+pi_gamma)
    
  }#End outer loop
  
  #Find average log likelihood
  log_lik <- log(lik/sim)

  #Return negative log likelihood
  ret <- -sum(log_lik)
  ret

}

```
