---
title: Meta-Analsis for the Binomial Model with Normal Random Effects
subtitle: Heirarchical and Empirical Bayesian Approaches
author: Sophie Huebler
format: revealjs
embed-resources: true
---



# Section 1: Introduction to the Problem

## The Goal
1) Create a computationally efficient implementation of the gibbs-metropolis-hastings hybrid algorithm for a heirarchical bayesian approach to meta analysis.

2) Create a computationally efficient function to compute mariginal likelihood and optimize it to empirically estimate parameters for random effects for an empirical bayesian approach to meta analysis.


## Meta-Analysis {.scrollable .smaller}
We will use a classic meta-analysis case to motivate this problem. Our treatment effect of interest is the odds ratio of an event occurring between treatment and control groups, and there are 7 studies which have estimated this effect by recording the number of events and sample size in a treatment sample and a control sample. 


```{r}
#| label: some-code
library(MASS)
library(epiworldR)
library(tidyverse)
library(metafor)
```

```{r}
source("data/read_data.R")
```

```{r}
dat |>
  dplyr::select(trial:nic)|>
  kableExtra::kbl(digits = 2,
      caption = "Data",
      col.names = c("Trial",
                    "R_it", "N_it",
                    "R_ic", "N_ic"))|>
  kableExtra::kable_classic_2(html_font = "Cambria")
```

## Meta-Analysis {.scrollable .smaller}

Each study therefore has a log odds ratio $\theta_i$ which estimates the study specific treatment effect, and a standard error for that estimate. We assume that due to differing study protocol and random chance, each study is estimating a log odds ratio that is drawn from a normal distribution with center $\mu$ and variance $\tau^2$. We will call this the random effect population prior $\pi(\theta_i | \mu, \tau)$.

```{r}
dat |>
  dplyr::select(trial:v_i)|>
  kableExtra::kbl(digits = 2,
      caption = "Data",
      col.names = c("Trial",
                    "R_it", "N_it",
                    "R_ic", "N_ic",
                    "log(OR)", "Var(log(OR))"))|>
  kableExtra::kable_classic_2(html_font = "Cambria")
```


## The Model

$$
\begin{aligned}
&i \in i,…,k \hspace{0.5cm} and \hspace{0.5cm} j \in \{C,T\}\\
& Y_{ij}| \pi_{ij},n_{ij} \sim Bin(n_{ij}, \pi_{ij})\\
&logit \pi_i^C | \gamma_i, \theta_i = \gamma_i - \theta_i/2\\
&logit \pi_i^T | \gamma_i, \theta_i = \gamma_i + \theta_i/2\\
& logit(\frac{\pi_{iT}}{\pi_{iC}}) = \theta_i | \mu, \tau^2 \sim N(\mu, \tau^2)\\
& \mu \sim N(a_1, b_1 )\\
& 1/\tau^2\sim gamma(a_2, b_2)\end{aligned}
$$


## Heirarichal Estimation

From the model, we can easily write out the joint distribution and the full conditionals for $\mu$, $\tau^2$, $\bf{\theta}$, and $\bf{\gamma}$ (I leave this as an exercise to the listener). This allows us to use the gibbs sampling procedure to iteratively update posteriors. It is however important to note that the thetas and gammas do not have closed form full conditionals. Therefore we must also incorporate a metropolis hastings component wise updating procedure. 

## Empirical Estimation

Instead of iteratively updating the $\mu$ and $\tau^2$ variables, we can estimate them from the marginal likelihood using an MLE approach after integrating out all other parameters, thus reducing the problem to a single level model.



# Section 2: Solution Plan

## Heirarchical Bayesian Approach

Full conditionals:
- Write in both R and c++
- Use microbenchmark to compare efficiency

Wrapper for the full algorithm:
- Rewrite with data table so that storage and manipulation is easier
- Pre-allocate memory

## Empirical Bayesian Approach

Marignal likelihood:
- Use parallel programming for the R implementation that currently simulates average likelihood of the marginal likelihood
- Write the full likelihood function in C++ and then use GNU scientific library for numeric integration


# Section 3: Preliminary Results

## Functions
Currently all posterior functions are written in R.  

## Marginal Likelihood

$$
\begin{aligned}
&\int_{\boldsymbol{\theta}} \int_{\boldsymbol{\gamma}} {n_i^T \choose r_i^T}{n_i^C \choose r_i^C} \frac{e^{(\gamma_i-\theta_i/2)^{r_i^C}}e^{(\gamma_i+\theta_i/2)^{r_i^T}}}{(1+ e^{\gamma_i-\theta_i/2})^{r_i^C}(1+e^{\gamma_i+\theta_i/2})^{r_i^T}}\frac{1}{\sqrt{2\pi 100}}e^{-\frac{1}{2*100}(\gamma_i-0)^2}\\
& \times \frac{1}{\sqrt{2\pi \tau^2}}e^{-\frac{1}{2*\tau^2}(\theta_i-\mu)^2}
\frac{0.001^{0.001}}{\Gamma(0.001)}(1/\tau^2)^{0.001-1}e^{\frac{-0.001}{\tau^2}} d\boldsymbol{\gamma}d\boldsymbol{\theta}
\end{aligned}
$$



